3D Regression Heat Map Analysis of Population Study Data

### 2015-06-24 draft.tex
revised with Steffis feedback

\design{
\begin{itemize}
	%%
  \item An overview visualization design based on feedback of epidemiological domain experts to support hypothesis-generation w.r.t. a target feature using regression models.
	%%
  \item Applying domain knowledge in the analysis using freely adjustable regression formulas.
	%%
  \item Selecting suitable metrics for representing and comparing regression models, as well as incorporating additional metrics for details on demand to assess the model quality.
	%%
  \item An openly available web-based system, which incorporates open formats to be applicable for data of other application domains.
	%%
\end{itemize}
}
% \begin{itemize}
% 	%%
% 	\item An overview visualization technique representing feature interactions using target features.
% 	%%
% 	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details on demand techniques for in-depth investigations of feature relationships.
% 	%%
% 	\item Freely adjustable regression formulas to provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
% 	%%
% 	\item The analysis of confounding features by providing comparison views between different formula results.
% 	%%
% 	\item The expansion and adaption of our method to any data with our open web-based system.
% 	%%
% \end{itemize}
%%
% \noindent The term `Regression Cube' has been used for the detailed analysis of one or a small number of regression models \cite{Ahmadi, Chan}.
% %%
% They extend 2D scatter plots to display detailed information about a regression model.
% %%
% In behalf of that, we declare our visualization technique for comparing and analyzing a large number of different regression models with different input features as \emph{Regression Cube}.
% %%
\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow, epidemiological data in general and the SHIP in particular.
%%
\subsection{Epidemiological Workflow} \label{sec:EpidemiologicalWorkflow}
%%
Epidemiological research is performed by experts from different academic disciplines, such as epidemiologists, physicians, statisticians, and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is to derive disease-specific risk factors by assessing epidemiological features with statistical methods.
%%
As described by Thew et al.~\cite{Thew2009} the epidemiological workflow is divided into these different steps:
\begin{enumerate}
	\item Clinicians/Epidemiologists derive hypotheses via clinical observations, experimental studies or literature research.
	%\item Clinicians make observations in the daily practice, which are translated into hypotheses.
	\item Epidemiologists compile a list of features depicting the hypothesis and include confounding features.
	\item Statisticians assess the association of the derived features w.r.t the investigated disease.
\end{enumerate}
%%
Relative risks can be determined if a statistical resilient association of features with a condition is extracted.
%%
Derived absolute risks indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is an epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS}, are used to analyze the data regarding the classical sequential epidemiological workflow.
%%
Hence, images are mostly used for the communication of results, rather than for providing insight into the data.
%%

@@

\subsection{Analysis Workflow} \label{sec:Workflow}
%%
% \begin{figure}[htb]
%  \centering
%  \includegraphics[width=1.0\linewidth]{figures/workflow}
%  \caption{
%  Different workflow types using the \emph{3D Regression Heat Map}.
%  %%
%  [\textbf{1.AF}] The workflow starts by declaring a formula to specify a hypothesis, or to use a predefined formula for a hypothesis-free analysis.
%  %%
%  [\textbf{2.SI}] The \emph{3D Regression Heat Map} is then visualized.
%  %%
%  The user has the option to either adjust the formula, adjust the transfer function or to derive details on demand on models.
%  %%
%  [\textbf{3.ZF}] Insights into the data yield either an adjustment of the current formula or a selection of a difference view.
%  %%
%  The latter is used to compare \emph{3D Regression Heat Maps}.
%  %%
%  [\textbf{4.DD}] Details about features using the 2D heat map representation yield insights and hypotheses about feature relations.
%  }
%   \label{fig:Workflow}
% \end{figure}
%%
\noindent Our \emph{3D Regression Heat Map} model is well suited for different workflow analysis techniques, based on the Visual Analytics (VA) Mantra of Keim et al.~\cite{Keim}:

@@

\textbf{Details on demand} [\textbf{4.DD}]. Precise information about the individual regression models (coefficients, associated confidence intervals and p-values) can be retrieved based on the data point representatives (e.g. in a hover modal on a currently selected data point).
\\\\
We use the squared bracket abbreviation for each step to denote the affiliation to the system design section later on.
%%
%As shown in Fig.~\ref{fig:Workflow}, the workflow is highly iterative.
The workflow is highly iterative.
%%
Observations in the 2D heat map or simply the CFS-based features can trigger new analyses by adjusting the underlying regression formulas.
%%
This can be carried out either to refine the current formula based on observations, or to create a new \emph{3D Regression Heat Map} for a difference view.
%%
\paragraph{Hypothesis-free and hypothesis-based analysis.}
%%
\design{
%%
Early analysis sessions yielded two approaches of analyzing the data.
%%
The classic approach is \emph{hypothesis-based}, where the expert already knows the data and potential associations (e.g. reproducing knowledge about hepatic steatosis risk factors based on known risk factors).
%%
The \emph{hypothesis-free} analysis allows to derive new insights, such as identifying confounding features or potential targets (e.g. deriving risk factors for breast cancer associated features).
%%
}
%%
%Input formulas reflect \emph{hypotheses} about the data.
\emph{Hypotheses} about the data are reflected using input formulas.
%%
Using the operators, dynamic variables and dataset features, many different assumptions can be expressed.
%%
To support the \emph{hypothesis-free} analysis, we provide a default formula:

@@

\paragraph{3D prism as data mini-map.}
\design{
%%
In early prototype versions, the 3D prism acted as starting point for the data analysis without the implementation of a separate 2D view.
%%
Slices were shown using cutaway planes.
%%
This approach was not popular among epidemiologists, because the complexity of the visualization overwhelmed them. %did not like the combination of 3D interaction and slicing with representatives.
%%
}
%%
The \emph{3D Regression Heat Map} representation \design{was redesigned to} act as an overview over the whole data set.
%%
%Its purpose is not to derive detailed information about data points.
%%
It serves as a function similar to a mini-map, guiding the attention to points of interest in the data.
%%
It also gives context information about adjacent data values when using the 2D heat map.
%%
\design{The distinction between overview and details on demand using two different representations was well received with our domain experts.}
%%
The displayed prism shows values above the matrix diagonal.
%%
For formulas with a dynamic target feature (e.g. exploratory analysis using $Z \sim X + Y$), the color encodes the absolute regression metric values (Fig.~\ref{fig:System}b).
%%
Applying this strategy to a formula containing a static target (e.g. $Cancer \sim X + Y + Z$) yields many occlusions, since the CFS algorithm creates the same feature space for every slice.
%%
For such formulas, the 3D view encodes every data element as absolute difference between its regression metric values and the global mean along the z-axis.
%%
This highlights slices with unusually low or high results (Fig.~\ref{fig:Application}).
%%
\magenta{Variables are ordered the same way in the 2D and 3D heat map to preserve the mental model and make them visually analogous.}

@@

%%
\paragraph{Attention steering is crucial.}
%%
Important events have to be highlighted in overview visualizations to direct the user's attention to interesting parts of the data.
%%
Poor guidance potentially leads to overlooked relationships.
%%
We found the 3D heat map as supporting mini-map visualizations most useful for this purpose, e.g., for highlighting differences rather than displaying absolute values (Fig.~\ref{fig:Application}).
%%
\section{Summary and Outlook}
%%
We presented a technique for knowledge discovery in population study data sets with user-defined target features.
%%
Dimension reduction using the target restricts the analysis to the most important features.
%%
\emph{Hypothesis-free} analysis employs default regression models.
%%
Modeling expert knowledge using regression formulas allows for a \emph{hypothesis-based} investigation.
%%
A \emph{3D Regression Heat Map} allows to assess hotspots in the analysis by abstracting regression models using a quality-of-fit measure.
%%
These can then be analyzed further using the 2D plot for each 3D heat map slice.
%%
Details on demand for each model allow for a detailed assessment of regression models.
%%
We successfully applied the approach to find correlations in a hepatic steatosis as well as a breast density data set.
%%
The method was well received by our clinical partners, triggering detailed investigations of the findings.

>>>>

\design{
\begin{itemize}
	%%
  \item An overview visualization design based on feedback of epidemiological domain experts to support hypothesis-generation w.r.t. a target feature using regression models.
	%%
  \item Applying domain knowledge in the analysis using freely adjustable regression formulas.
	%%
  \item Selecting suitable metrics for representing and comparing regression models, as well as incorporating additional metrics for details-on-demand to assess the model quality.
	%%
  \item An openly available web-based system, which incorporates open formats to be applicable for data of other application domains.
	%%
\end{itemize}
}
% \begin{itemize}
% 	%%
% 	\item An overview visualization technique representing feature interactions using target features.
% 	%%
% 	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details-on-demand techniques for in-depth investigations of feature relationships.
% 	%%
% 	\item Freely adjustable regression formulas to provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
% 	%%
% 	\item The analysis of confounding features by providing comparison views between different formula results.
% 	%%
% 	\item The expansion and adaption of our method to any data with our open web-based system.
% 	%%
% \end{itemize}
%%
% \noindent The term `Regression Cube' has been used for the detailed analysis of one or a small number of regression models \cite{Ahmadi, Chan}.
% %%
% They extend 2D scatter plots to display detailed information about a regression model.
% %%
% In behalf of that, we declare our visualization technique for comparing and analyzing a large number of different regression models with different input features as \emph{Regression Cube}.
% %%
\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow, epidemiological data in general and the SHIP in particular.
%%
\subsection{Epidemiological Workflow} \label{sec:EpidemiologicalWorkflow}
%%
Epidemiological research is performed by experts from different academic disciplines, such as epidemiologists, physicians, statisticians, and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is to derive disease-specific risk factors by assessing epidemiological features with statistical methods.
%%
As described by Thew et al.~\cite{Thew2009} the epidemiological workflow is divided into these different steps:
\begin{enumerate}
	\item Clinicians/Epidemiologists derive hypotheses via clinical observations, experimental studies or literature research.
	%\item Clinicians make observations in the daily practice, which are translated into hypotheses.
	\item Epidemiologists compile a list of features depicting the hypothesis and include confounding features.
	\item Statisticians assess the association of the derived features w.r.t the investigated disease.
\end{enumerate}
%%
Relative risks can be determined if a statistical resilient association of features with a condition is extracted.
%%
Derived absolute risks indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is an epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS}, are used to analyze the data regarding the classical sequential epidemiological workflow.
%%
Hence, images are mostly used for the communication of results, rather than for providing insight into the data.
%%

@@

\subsection{Analysis Workflow} \label{sec:Workflow}
%%
% \begin{figure}[htb]
%  \centering
%  \includegraphics[width=1.0\linewidth]{figures/workflow}
%  \caption{
%  Different workflow types using the \emph{3D Regression Heat Map}.
%  %%
%  [\textbf{1.AF}] The workflow starts by declaring a formula to specify a hypothesis, or to use a predefined formula for a hypothesis-free analysis.
%  %%
%  [\textbf{2.SI}] The \emph{3D Regression Heat Map} is then visualized.
%  %%
%  The user has the option to either adjust the formula, adjust the transfer function or to derive details-on-demand on models.
%  %%
%  [\textbf{3.ZF}] Insights into the data yield either an adjustment of the current formula or a selection of a difference view.
%  %%
%  The latter is used to compare \emph{3D Regression Heat Maps}.
%  %%
%  [\textbf{4.DD}] Details about features using the 2D heat map representation yield insights and hypotheses about feature relations.
%  }
%   \label{fig:Workflow}
% \end{figure}
%%
\noindent Our \emph{3D Regression Heat Map} model is well suited for different workflow analysis techniques, based on the Visual Analytics (VA) Mantra of Keim et al.~\cite{Keim}:

@@

\textbf{Details-on-demand} [\textbf{4.DD}]. Precise information about the individual regression models (coefficients, associated confidence intervals and p-values) can be retrieved based on the data point representatives (e.g. in a hover modal on a currently selected data point).
\\\\
We use the squared bracket abbreviation for each step to denote the affiliation to the system design section later on.
%%
%As shown in Fig.~\ref{fig:Workflow}, the workflow is highly iterative.
The workflow is highly iterative.
%%
Observations in the 2D heat map or simply the CFS-based features can trigger new analyses by adjusting the underlying regression formulas.
%%
This can be carried out either to refine the current formula based on observations, or to create a new \emph{3D Regression Heat Map} for a difference view.
%%
\paragraph{Hypothesis-free and hypothesis-based analysis.}
%%
\design{
%%
Early analysis sessions yielded two approaches of analyzing the data.
%%
The classic approach is \emph{hypothesis-based}, where the expert already knows the data and potential associations (e.g. reproducing knowledge about hepatic steatosis risk factors based on known risk factors).
%%
The \emph{hypothesis-free} analysis allows to derive new insights, such as identifying confounding features or potential targets (e.g. deriving risk factors for breast cancer associated features).
%%
}
%%
%Input formulas reflect \emph{hypotheses} about the data.
\emph{Hypotheses} about the data are reflected using input formulas.
%%
Using the operators, dynamic variables and dataset features, many different assumptions can be expressed.
%%
To support the \emph{hypothesis-free} analysis, we provide a default formula:

@@

\paragraph{3D prism as data mini-map.}
\design{
%%
In early prototype versions, the 3D prism acted as starting point for the data analysis without the implementation of a separate 2D view.
%%
Slices were shown using cutaway planes.
%%
This approach was not popular among epidemiologists, because the complexity of the visualization overwhelmed them. %did not like the combination of 3D interaction and slicing with representatives.
%%
}
%%
The \emph{3D Regression Heat Map} representation \design{was redesigned to} act as an overview over the whole data set.
%%
%Its purpose is not to derive detailed information about data points.
%%
It serves as a function similar to a mini-map, guiding the attention to points of interest in the data.
%%
It also gives context information about adjacent data values when using the 2D heat map.
%%
\design{The distinction between overview and details-on-demand using two different representations was well received with our domain experts.}
%%
The displayed prism shows values above the matrix diagonal.
%%
For formulas with a dynamic target feature (e.g. exploratory analysis using $Z \sim X + Y$), the color encodes the absolute regression metric values (Fig.~\ref{fig:System}b).
%%
Applying this strategy to a formula containing a static target (e.g. $Cancer \sim X + Y + Z$) yields many occlusions, since the CFS algorithm creates the same feature space for every slice.
%%
For such formulas, the 3D view encodes every data element as absolute difference between its regression metric values and the global mean along the z-axis.
%%
This highlights slices with unusually low or high results (Fig.~\ref{fig:Application}).
%%
\magenta{Variables are ordered the same way in the 2D and 3D heat map to preserve the mental model and make them visually analogous.}

@@

%%
\paragraph{Attention steering is crucial.}
%%
Important events have to be highlighted in overview visualizations to direct the user's attention to interesting parts of the data.
%%
Poor guidance potentially leads to overlooked relationships.
%%
We found the 3D heat map as supporting mini-map visualizations most useful for this purpose, e.g., for highlighting differences rather than displaying absolute values (Fig.~\ref{fig:Application}).
%%
\section{Summary and Outlook}
%%
We presented a technique for knowledge discovery in population study data sets with user-defined target features.
%%
Dimension reduction using the target restricts the analysis to the most important features.
%%
\emph{Hypothesis-free} analysis employs default regression models.
%%
Modeling expert knowledge using regression formulas allows for a \emph{hypothesis-based} investigation.
%%
A \emph{3D Regression Heat Map} allows to assess hotspots in the analysis by abstracting regression models using a quality-of-fit measure.
%%
These can then be analyzed further using the 2D plot for each 3D heat map slice.
%%
Details-on-demand for each model allow for a detailed assessment of regression models.
%%
We successfully applied the approach to find correlations in a hepatic steatosis as well as a breast density data set.
%%
The method was well received by our clinical partners, triggering detailed investigations of the findings.
